---
title: "Project_5-Prediction_of_Drinking_Water_Potability"
author: "Tiffany & Fahad"
knit:: (function(inputFile,
  encoding){
  out_dir <- '../output';
  rmarkdown::render(inputFile,
  encoding = encoding,
  output_file = file.path(dirname(inputFile),
  out_dir,
  'Week8Assignment.html')) })


output:
  html_document:
    theme: cosmo
    toc: true
  html_notebook: default
---

# Project #5: Drinking Water Potability

## Description
Context Access to safe drinking water is essential to health, a basic human right, and a component of effective policy for health protection. This is important as a health and development issue at a national, regional, and local level. In some regions, it has been shown that investments in water supply and sanitation can yield a net economic benefit, since the reductions in adverse health effects and health care costs outweigh the costs of undertaking the interventions.

## Null Hypothesis
There is no relationship between optimal levels of chemical components contained in treated water and potability of consumed water.

See if you can find any other trends in the data to predict potability of consumed water or find any clear indications of relationships between the predictors.

## Content
The drinking_water_potability.csv file contains water quality metrics for 3276 different water bodies.
  1. pH value: PH is an important parameter in evaluating the acid-base balance of water. It is also the indicator of the acidic or alkaline condition of water status. WHO has recommended the maximum permissible limit of pH from 6.5 to 8.5. The current investigation ranges were 6.52–6.83 which are in the range of WHO standards.
  2. Hardness: Hardness is mainly caused by calcium and magnesium salts. These salts are dissolved from geologic deposits through which water travels. The length of time water is in contact with hardness-producing material helps determine how much hardness there is in raw water. Hardness was originally defined as the capacity of water to precipitate soap caused by Calcium and Magnesium.
  3. Solids (Total dissolved solids - TDS): Water has the ability to dissolve a wide range of inorganic and some organic minerals or salts such as potassium, calcium, sodium, bicarbonates, chlorides, magnesium, sulfates, etc. These minerals produced an unwanted taste and diluted color in the appearance of water. This is the important parameter for the use of water. The water with a high TDS value indicates that water is highly mineralized. The desirable limit for TDS is 500 mg/l and the maximum limit is 1000 mg/l which is prescribed for drinking purposes.
  4. Chloramines: Chlorine and chloramine are the major disinfectants used in public water systems. Chloramines are most commonly formed when ammonia is added to chlorine to treat drinking water. Chlorine levels up to 4 milligrams per liter (mg/L or 4 parts per million (ppm)) are considered safe in drinking water.
  5. Sulfate: Sulfates are naturally occurring substances that are found in minerals, soil, and rocks. They are present in ambient air, groundwater, plants, and food. The principal commercial use of sulfate is in the chemical industry. Sulfate concentration in seawater is about 2,700 milligrams per liter (mg/L). It ranges from 3 to 30 mg/L in most freshwater supplies, although much higher concentrations (1000 mg/L) are found in some geographic locations.
  6. Conductivity: Pure water is not a good conductor of electric current rather’s a good insulator. An increase in ions concentration enhances the electrical conductivity of water. Generally, the amount of dissolved solids in water determines electrical conductivity. Electrical conductivity (EC) actually measures the ionic process of a solution that enables it to transmit current. According to WHO standards, EC value should not exceed 400 μS/cm.
  7. Organic_carbon: Total Organic Carbon (TOC) in source waters comes from decaying natural organic matter (NOM) as well as synthetic sources. TOC is a measure of the total amount of carbon in organic compounds in pure water. According to US EPA < 2 mg/L as TOC in treated / drinking water, and < 4 mg/Lit in source water which is use for treatment.
  8. Trihalomethanes: THMs are chemicals that may be found in water treated with chlorine. The concentration of THMs in drinking water varies according to the level of organic material in the water, the amount of chlorine required to treat the water, and the temperature of the water that is being treated. THM levels up to 80 ppm are considered safe in drinking water.
  9. Turbidity: The turbidity of water depends on the quantity of solid matter present in the suspended state. It is a measure of the light-emitting properties of water and the test is used to indicate the quality of waste discharge with respect to the colloidal matter. The mean turbidity value obtained for Wondo Genet Campus (0.98 NTU) is lower than the WHO recommended value of 5.00 NTU.
  10. Potability: Indicates if water is safe for human consumption where 1 means Potable and 0 means Not potable.

## Code

### Set up
Libraries:
```{r Libraries}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(DT)
library(knitr)
library(kableExtra)
library(psych)
library(gridExtra)
library(grid)
library(lattice)
library(caret)
library(car)
library(pastecs)
library(psych)
library(wesanderson)
library(e1071)
library(boot)
library(ggm)
library(Hmisc)
library(polycor)
library(ppcor)
library(corrplot)
library(formattable)
library(ggbeeswarm)
library(gmodels)
```

```{r Working Directory}
getwd()
```

### Data Analysis
Read in Dataset:
```{r Dataset}
dw.potability <- read.csv("./Data/drinking_water_potability.csv")
head(dw.potability)
kable(head(dw.potability), booktabs =T) %>% 
  kable_styling(latex_options ="striped")

str(dw.potability)
```

Check for NA values:
```{r NA Check}
ph_na <- is.na(dw.potability$ph)
table(ph_na)

Hardness_na <- is.na(dw.potability$Hardness)
table(Hardness_na)

Solids_na <- is.na(dw.potability$Solids)
table(Solids_na)

Chloramines_na <- is.na(dw.potability$Chloramines)
table(Chloramines_na)

Sulfate_na <- is.na(dw.potability$Sulfate)
table(Sulfate_na)

Conductivity_na <- is.na(dw.potability$Conductivity)
table(Conductivity_na)

carbon_na <- is.na(dw.potability$Organic_carbon)
table(carbon_na)

Trihalomethanes_na <- is.na(dw.potability$Trihalomethanes)
table(Trihalomethanes_na)

Turbidity_na <- is.na(dw.potability$Turbidity)
table(Turbidity_na)

#[@tc] Variables that need to be cleaned: ph, Sulfate, Trihalomethanes
```

Omit NA values:
```{r Omit NA}
#Variables where there are NA values: ph, Sulfate, Trihalomethanes
na_omit_dw <- na.omit(dw.potability)
head(na_omit_dw, 10)
```

Categorical Variable: Potability
Potability: Indicates if water is safe for human consumption where 1 means Potable and 0 means Not potable.
```{r Categorical Variable}
dw.potability$potability_cat[dw.potability$Potability == 1] = "Potable"
dw.potability$potability_cat[dw.potability$Potability == 0] = "Not Potable"

dw.potability$potability_cat <- as.factor(dw.potability$potability_cat)

str(dw.potability)
```
```{r NA Omit Categorical Variable}
na_omit_dw$potability_cat[na_omit_dw$Potability == 1] = "Potable"
na_omit_dw$potability_cat[na_omit_dw$Potability == 0] = "Not Potable"

na_omit_dw$potability_cat <- as.factor(na_omit_dw$potability_cat)

str(na_omit_dw)
```

Numerical Variable Summaries:
```{r Numerical Variable}
summary(na_omit_dw$ph)
summary(na_omit_dw$Hardness)
summary(na_omit_dw$Solids)
summary(na_omit_dw$Chloramines)
summary(na_omit_dw$Sulfate)
summary(na_omit_dw$Conductivity)
summary(na_omit_dw$Organic_carbon)
summary(na_omit_dw$Trihalomethanes)
summary(na_omit_dw$Turbidity)

str(na_omit_dw)
```

Frequency Diagrams:
```{r Frequency}
#Frequency diagrams for numerical variables, grouped by categorical variable potability_cat
ph_hist <- ggplot(data = dw.potability,
                  aes(x = ph,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 0.5) +
  labs(title = "ph Histogram")
ph_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$ph,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$ph,
                                    na.rm = TRUE)),
                size = 1)

Hardness_hist <- ggplot(data = dw.potability,
                  aes(x = Hardness,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 10) +
  labs(title = "Hardness Histogram")
Hardness_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Hardness,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Hardness,
                                    na.rm = TRUE)),
                size = 1)

Solids_hist <- ggplot(data = dw.potability,
                  aes(x = Solids,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 1500) + 
  labs(title = "Solids Histogram")
Solids_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Solids,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Solids,
                                    na.rm = TRUE)),
                size = 1)

Chloramines_hist <- ggplot(data = dw.potability,
                  aes(x = Chloramines,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 0.5) +
  labs(title = "Chrolamines Histogram")
Chloramines_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Chloramines,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Chloramines,
                                    na.rm = TRUE)),
                size = 1)

Sulfate_hist <- ggplot(data = dw.potability,
                  aes(x = Sulfate,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 10) + 
  labs(title = "Sulfate Histogram")
Sulfate_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Sulfate,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Sulfate,
                                    na.rm = TRUE)),
                size = 1)

Conductivity_hist <- ggplot(data = dw.potability,
                  aes(x = Conductivity,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 20) +
  labs(title = "Conductivity Histogram")
Conductivity_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Conductivity,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Conductivity,
                                    na.rm = TRUE)),
                size = 1)

Carbon_hist <- ggplot(data = dw.potability,
                  aes(x = Organic_carbon,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 1) +
  labs(title = "Carbon Histogram")
Carbon_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Organic_carbon,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Organic_carbon,
                                    na.rm = TRUE)),
                size = 1)

Trihalomethanes_hist <- ggplot(data = dw.potability,
                  aes(x = Trihalomethanes,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 5) + 
  labs(title = "Trihalomethanes Histogram")
Trihalomethanes_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Trihalomethanes,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Trihalomethanes,
                                    na.rm = TRUE)),
                size = 1)

Turbidity_hist <- ggplot(data = dw.potability,
                  aes(x = Turbidity,
                      fill = potability_cat),
                  na.rm = TRUE) +
  geom_histogram(data = dw.potability,
                 aes(y = ..density..),
                 na.rm = TRUE,
                 binwidth = 0.25) + 
  labs(title = "Turbidity Histogram")
Turbidity_hist +
  stat_function(fun = dnorm,
                args = list(mean = mean(dw.potability$Turbidity,
                                        na.rm = TRUE),
                            sd = sd(dw.potability$Turbidity,
                                    na.rm = TRUE)),
                size = 1)
```

Potability Bar Chart:
```{r Bar Chart}
potability_bar <- ggplot(data = dw.potability,
                         aes(x = potability_cat,
                             fill = potability_cat)) + 
  geom_bar() +
  labs(title = "Potability Bar Chart")
potability_bar
```

```{r Class Imbalance}
table(dw.potability$potability_cat)

CI_potability <-  (1998 - 1278)/(1998 + 1278)
CI_potability
```

### Exploratory Data Analysis
#### Null Hypothesis
There is no relationship between optimal levels of chemical components contained in treated water and potability of consumed water.

See if you can find any other trends in the data to predict potability of consumed water or find any clear indications of relationships between the predictors.

Numerical Variable Subset:
```{r Numerical Variable Subset}
#Numerical Variables: ph, Hardness, Solids, Chloramines, SUlfate, Conductivity, Organic_carbon, Trihalomethanes, Turbidity
potability_num <- na_omit_dw[ ,
                                 c(1, 2, 3, 4, 5, 6, 7, 8, 9)]
head(potability_num, 10)
```

#### Variable Comparisons
Correlations:
```{r Correlation Matrix}
M <- correlations(x = na_omit_dw[ , 
                                     c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)])
M
```

```{r Correlogram}
dw_corr <- corrplot(M,
                    method = "color",
                    title = "Drinking Water Potability Correlogram")
dw_corr

# [@tc] The most statistically significant relationship seems to be between Sulfate and Solids levels.
# Correlation: -0.164142092
```

Scatterplots, grouped by Potability
```{r Grouped Scatterplots}
# ph vs. Hardness
# Correlation: 0.0842915812
ph_Hardness <- ggplot(data = dw.potability,
                      aes(x = Hardness,
                          y = ph,
                          color = potability_cat),
                      na.rm = TRUE) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm) +
  labs(title = "ph vs. Hardness Scatterplot")
ph_Hardness

# ph vs. Solids
# Correlation: -0.073568502
ph_Solids <- ggplot(data = dw.potability,
                      aes(x = Solids,
                          y = ph,
                          color = potability_cat),
                      na.rm = TRUE) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm) +
  labs(title = "ph vs. Solids Scatterplot")
ph_Solids

# Hardness vs. Solids
# Correlation: -0.063815119
Hardness_Solids <- ggplot(data = dw.potability,
                      aes(x = Solids,
                          y = Hardness,
                          color = potability_cat),
                      na.rm = TRUE) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm) +
  labs(title = "Hardness vs. Solids Scatterplot")
Hardness_Solids

# Trihalomethanes vs. Solids
# Correlation: -0.054720374
Trihalomethanes_Solids <- ggplot(data = dw.potability,
                      aes(x = Solids,
                          y = Trihalomethanes,
                          color = potability_cat),
                      na.rm = TRUE) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm) +
  labs(title = "Trihalomethanes vs. Solids Scatterplot")
Trihalomethanes_Solids

# Sulfate vs. Solids
# Correlation: -0.164142092
Sulfate_Solids <- ggplot(data = dw.potability,
                      aes(x = Solids,
                          y = Sulfate,
                          color = potability_cat),
                      na.rm = TRUE) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm) +
  labs(title = "Sulfate vs. Solids Scatterplot")
Sulfate_Solids

# [@tc] It appears that the linear models in the Sulfate vs. Solids, ph vs. Hardness and ph vs. Solids scatterplots display very different gradients when grouped by Not Potable and Potable

# [@tc] The Sulfate vs. Solids plot seems to show that there is almost no relationship between Sulfate and Solids levels in non-potable water, but a visible negative relationship between Sulfate and Solids levels in potable water.
```
